---
title: "results"
author: "Jeremy Forbes"
date: "19/03/2019"
output: html_document
---

```{r}
library(spdep)
library(tidyverse)
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/model_df.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_16.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_13.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_10.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_07.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_04.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_01.rda")

fmod16 <- errorsarlm(LNP_Percent ~ ., 
  data=(model_df %>% filter(year == "2016") %>% dplyr::select(-c(year, DivisionNm))),
  sp_weights_16, etype="error", method="eigen", interval=c(-1,0.999))

fmod13 <- errorsarlm(LNP_Percent ~ ., 
  data=(model_df %>% filter(year == "2013") %>% dplyr::select(-c(year, DivisionNm))),
  sp_weights_13, etype="error", method="eigen", interval=c(-1,0.999))

fmod10 <- errorsarlm(LNP_Percent ~ ., 
  data=(model_df %>% filter(year == "2010") %>% dplyr::select(-c(year, DivisionNm))),
  sp_weights_10, etype="error", method="eigen", interval=c(-1,0.999))

fmod07 <- errorsarlm(LNP_Percent ~ ., 
  data=(model_df %>% filter(year == "2007") %>% dplyr::select(-c(year, DivisionNm))),
  sp_weights_07, etype="error", method="eigen", interval=c(-1,0.999))

fmod04 <- errorsarlm(LNP_Percent ~ ., 
  data=(model_df %>% filter(year == "2004") %>% dplyr::select(-c(year, DivisionNm))),
  sp_weights_04, etype="error", method="eigen", interval=c(-1,0.999))

fmod01 <- errorsarlm(LNP_Percent ~ ., 
  data=(model_df %>% filter(year == "2001") %>% dplyr::select(-c(year, DivisionNm))),
  sp_weights_01, etype="error", method="eigen", interval=c(-1,0.999))

# Coefficients and confidence bands
coef_df <- bind_rows(
  data.frame(variable = fmod16$coefficients %>% names, estimate = fmod16$coefficients %>% unname, se = fmod16$rest.se %>% unname, p = summary(fmod16)$Coef[,4] %>% unname, year = 2016),
  data.frame(variable = fmod13$coefficients %>% names, estimate = fmod13$coefficients %>% unname, se = fmod13$rest.se %>% unname, p = summary(fmod13)$Coef[,4] %>% unname, year = 2013),
  data.frame(variable = fmod10$coefficients %>% names, estimate = fmod10$coefficients %>% unname, se = fmod10$rest.se %>% unname, p = summary(fmod10)$Coef[,4] %>% unname, year = 2010),
  data.frame(variable = fmod07$coefficients %>% names, estimate = fmod07$coefficients %>% unname, se = fmod07$rest.se %>% unname, p = summary(fmod07)$Coef[,4] %>% unname, year = 2007),
  data.frame(variable = fmod04$coefficients %>% names, estimate = fmod04$coefficients %>% unname, se = fmod04$rest.se %>% unname, p = summary(fmod04)$Coef[,4] %>% unname, year = 2004),
  data.frame(variable = fmod01$coefficients %>% names, estimate = fmod01$coefficients %>% unname, se = fmod01$rest.se %>% unname, p = summary(fmod01)$Coef[,4] %>% unname, year = 2001)
)
```

## Results

```{r}
#stargazer::stargazer(fmod16)
```

### Spatial correlation

The spatial autoregressive coefficient represents the extent to which an electorate is affected by the surrounding political climate. In the 2001, 2013 and 2016 elections $\rho$ is positive and significant, indicating that electoral voting outcomes were influenced by attitudes of neighbouring electorates, in the sense that two-party preference. However, in the 2004, 2007 and 2010 elections, $\rho$ is not significantly different from zero, meaning that neighbours did not influence the two-party preferred vote over and above the effect of electoral socio-demographics.

```{r}
coef_df %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(upper95 = estimate + 1.96*se, lower95 = estimate - 1.96*se) %>% 
  ggplot() +
  geom_point(aes(x = year, y = estimate, col = factor(p < 0.05)), size = 2) +
  geom_linerange(aes(x = year, ymin = lower95, ymax = upper95, col = factor(p < 0.05)), size = 1.5) + 
  geom_hline(aes(yintercept = 0), alpha = 0.5, size = 1) + 
  facet_wrap(~variable, scales = "free") +
  scale_color_manual(values = c("grey50", "black")) +
  guides(col = F) +
  theme(axis.text = element_text(size = 8), axis.text.x = element_text(angle = 60, hjust=1)) +
  scale_x_continuous(breaks = c(2001, 2004, 2007, 2010, 2013, 2016))
```


### Persistent effects
Defacto - Labor
DiffAddress - Liberal
Extractive - Liberal
Indigenous - Liberal
Admin - Liberal

```{r, include = F}
coef_df %>% 
  filter(variable %in% c("DeFacto", "DiffAddress", "Extractive", "Indigenous", "ManagerAdminClericalSales")) %>% 
  mutate(upper95 = estimate + 1.96*se, lower95 = estimate - 1.96*se) %>% 
  ggplot() +
  geom_point(aes(x = year, y = estimate), size = 3) +
  geom_linerange(aes(x = year, ymin = lower95, ymax = upper95), size = 1.5) + 
  geom_hline(aes(yintercept = 0), alpha = 0.5, size = 1) + 
  facet_wrap(~variable) +
  scale_color_manual(values = c("grey50", "black"))
```


### Changing effects
Unemployment - Labor then none
FamilyHouseSize - Labor 2004-2010
Incomes - Liberal in 2016
MedianAge - Liberal 2004 and 2016

```{r, include = F}
coef_df %>% 
  filter(variable %in% c("Unemployment", "FamHouseSize", "Incomes", "MedianAge")) %>% 
  mutate(upper95 = estimate + 1.96*se, lower95 = estimate - 1.96*se) %>% 
  ggplot() +
  geom_point(aes(x = year, y = estimate, col = factor(p < 0.05)), size = 3) +
  geom_linerange(aes(x = year, ymin = lower95, ymax = upper95, col = factor(p < 0.05)), size = 1.5) + 
  geom_hline(aes(yintercept = 0), alpha = 0.5, size = 1) + 
  facet_wrap(~variable) +
  scale_color_manual(values = c("grey50", "black")) +
  guides(col = F)
```

### Alternative: discuss each election individually/over time

Over the six elections, it is apparent that five socio-demographic characterstics have been strongly associated with support for one of the parties in all elections, whilst others are influential in some, and others not at all. The liberal party appeals to XXX. Labor party XXX. In 2001, ....


### Diagnostics
Goodness of fit?

Residuals: GLS residuals should be normal (QQ-plot)
- fmod16$residuals

Outliers:
-- Leverage: Hat values (X)
-- Influence: Cook's distance

#### Electorates that are very different to their neighbours
- Residuals vs weighted neighbours residuals
- Socio-demographics that are quite different?



$${\bf r_j} = {\bf \hat{u}} + \hat\beta_j{\boldsymbol X_j}$$
where

$${\bf \hat{u}} = ({\bf I_n}-\hat\rho \cdot {\bf W}) {\bf  \hat v}$$

where 

$${\bf  \hat v} \sim N({\bf 0}, \sigma^2 \cdot {\bf I_n}) $$